---
title: A Bayesian spatial and temporal modeling approach to mapping geographic variation in infant mortality rates for canary counties with R-INLA
author: "Bello, L., Santana, A., Saavedra, P."
date: "12/06/2022"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
library(openxlsx)
library(tidyverse)
library(INLA)
library(janitor)
library(flextable)
library(xtable)
# Libraries for maps:
library(rgdal)
library(RColorBrewer)
library(ggspatial)
library(sf)
library(openxlsx)
```


### Data reading and definition of new data.frames
```{r}
# Original data
db <- read.xlsx("InfantMortality.xlsx")
n <- nrow(db) # number of counties

# Extract only county, births, deaths and deprivation index
# in periods 1994-2003 and 2004-2016
dt0 <- db %>% 
  select(Comarca,B_03,D_03,IP_03,B_16,D_16,IP_16) 

# Save in long format
dt <- dt0%>%
  mutate(County=1:n) %>% 
  pivot_longer(B_03:IP_16,names_to = c(".value","Period"),
               names_pattern = "(.)_(.)") %>% 
  rename(CountyName=Comarca,Birth=B,Death=D,DI=P) %>% 
  mutate(Period=as.numeric(Period),
         R_obs=1000*Death/Birth,
         County.Period = County*Period) %>% 
  arrange(Period,County)

# Global rates
dt %>% 
  group_by(Period) %>% 
  summarize(Birth=sum(Birth),
            Death=sum(Death)) %>% 
  mutate(R=1000*Death/Birth)
```

### Table 1: Births, deaths and deprivation index by county and period
```{r}
tb1 <- dt0 %>% mutate(` `="") %>% 
  select(1:4,8,5:7) %>% 
  adorn_totals("row", fill="",...=c(2,3,6,7))
# View table:
tb1 %>% flextable() %>% autofit()
# Save table as latex file:
print(xtable(tb1, type = "latex"),include.rownames=FALSE, file = "table1.tex")
```

### Figure 1: Deprivation Index Evolution
```{r}
ggplot(db,aes(x=IP_03,y=IP_16))+
  geom_point()+
  geom_abline()+
  labs(x="Deprivation index 1994 - 2003",y="Deprivation index 2004-2016") +
  theme_bw()
ggsave("figure1.eps",height=119,width=119,units="mm",dpi=1200)
```

### H-matrix: neighbourhood matrix for spatial model
```{r}
dh <- read.xlsx("MortalityRates.xlsx",2)
h <- as.matrix(dh[,-1])
attr(h,"dimnames")=NULL

# Check if the matrix is indeed symmetric:
con=0
for(i in 1:27)
  for(j in 1:27)
    if(h[i,j]!=h[j,i]) con=con+1
con

# Convert to inla format
H <- inla.graph2matrix(h)

# Visualize the neighbourhood matrix
lc <- db$Comarca
par(mar=c(7,6.3,4,1)) 
fields::image.plot(h,xaxt="n",yaxt="n",col=c("white","black")) 
mtext(text=lc, side=1,line=0.4,at=seq(1/27,1,1/27), las=2, cex=.6) 
mtext(text=lc, side=2,line=0.3,at=seq(1/27,1,1/27), las=1, cex=.6) 
```

### Table 2. Posterior distribution for the fixed parameters
```{r}
formula <- Death ~ 1 + f(County, model="bym",graph=H,constr=TRUE) +
  f(County.Period,model="iid",prior="normal",param=c(-0.4,1)) + DI

# Inla model fitting:
mp <- inla(formula, family="poisson", data=dt, offset=log(Birth),
           control.fixed=list(mean.intercept=0, prec.intercept=0.00001),
           control.predictor=list(compute=TRUE),
           control.compute=list(dic=TRUE,cpo=TRUE))

# summary(mp)
# mp$summary.fixed
a <- mp$summary.fixed[,c(4,3,5)]
b <- -mp$summary.random$County.Period[1,c(5,6,4)]
tp <- data.frame(rbind(a,b))
rownames(tp) <- NULL
tb2 <- tp %>% mutate(Parameter=c("Intercept","Deprivation index","Period"),
              `Estimate (95% Credibility Interval)`= 
                sprintf("%.3f (%.3f ; %.3f)",tp[[1]],tp[[2]],tp[[3]])) %>% 
  select(-(1:3))

# View table:
tb2 %>% flextable() %>% autofit()
# Save table as latex file:
print(xtable(tb2, type = "latex"),include.rownames=FALSE, file = "table2.tex")

# mean(mp$summary.random$County.Period$mean[2:28])
```

### Figure 2: Goodness of fit
```{r}
data.frame(Death=dt$Death,
           lambda_mu = mp$summary.fitted.values$mean) %>% 
  ggplot(aes(Death,lambda_mu)) +
  geom_point() +
  geom_abline() +
  theme_bw() +
  labs(x="Deaths count",y="Fitted expected number of deaths")
ggsave("figure2.eps",height=119,width=119,units="mm",dpi=1200)
```

### Figure 3: Overall rate ratio
```{r}
##Overall Rate Ratio:
(grr <- exp(b))

data.frame(mp$marginals.random$County.Period$index.1) %>% 
  mutate(xt=exp(-x),
         yt=exp(x)*y) %>% 
  ggplot(aes(xt,yt)) +
  geom_line() +
  labs(x="Overall rate ratio",y="Posterior probability density") +
  xlim(0.35,0.9) +
  geom_vline(xintercept = as.numeric(grr[1]),linetype="longdash") +
  theme_bw() +
  coord_cartesian(ylim=c(0,5), expand = FALSE) 
ggsave("figure3.eps",height=70,width=119,units="mm",dpi=1200)

```


### Table 3. Estimation of lambda
```{r}
### Mortality rates
lambdaFit_03 <- mp$summary.fitted.values[1:n,]
R_fit_03 <- 10^3*lambdaFit_03/db$B_03
lambdaFit_16 <- mp$summary.fitted.values[(n+1):(2*n),]
R_fit_16 <- 10^3*lambdaFit_16/db$B_16
tb03 <- sprintf("%.2f (%.2f ; %.2f)",R_fit_03[,4],R_fit_03[,3],R_fit_03[,5])
tb16 <- sprintf("%.2f (%.2f ; %.2f)",R_fit_16[,4],R_fit_16[,3],R_fit_16[,5])

### Ratio rates of mortality rates (beta + d_{i})
mat <- cbind(rep(-1,n),diag(n))
lci <- inla.make.lincombs(County.Period=mat)
mpi <- inla(formula,family="poisson",data=dt,offset=log(Birth),lincomb=lci,
           control.predictor=list(compute=TRUE),control.compute=list(dic=TRUE,cpo=TRUE))
lpi <- mpi$summary.lincomb.derived
rr <- exp(lpi[,4:6])
tbRR <- sprintf("%.2f (%.2f ; %.2f)",rr[,2],rr[,1],rr[,3]) 

tb3 <- data.frame(lc,tb03,tb16,tbRR)
names(tb3) <- c("County","1994 - 2003","2004 - 2016","Ratio rates")

# View table:
tb3 %>% flextable() %>% autofit()
# Save table as latex file:
print(xtable(tb3, type = "latex"),include.rownames=FALSE, file = "table3.tex")

### Comprobación
R_fit_16$mean/R_fit_03$mean
```

### Figure 4: Map
```{r}
# Geographic map was originally downloaded from
# https://datos.canarias.es/catalogos/estadisticas/dataset/comarcas-de-canarias-delimitaciones-territoriales-para-fines-estadisticos

# Data for the map:
r=sum(dt$Death)/sum(dt$Birth)
E=r*dt$Birth
Rho_03 <- lambdaFit_03/E[1:n]
Rho_16 <- lambdaFit_16/E[(n+1):(2*n)]
dr <- data.frame(County=lc,Rho_03$mean,Rho_16$mean) # rate ratios

comarcasCanarias <- readOGR("mapaComarcas/comarcas_20170101.shp")
etiquetasComarcas <- read.xlsx("CountyLabels.xlsx")
canarias <- st_as_sf(comarcasCanarias)
rates <- dr %>% 
  full_join(etiquetasComarcas) %>% 
  pivot_longer(Rho_03.mean:Rho_16.mean, names_to = "Year", values_to="Rho0") %>% 
  mutate(Year=factor(Year,levels=c("Rho_03.mean","Rho_16.mean"),
                     labels=c("Period 1994-2003","Period 2004-2016")),
         Rho=cut(Rho0,breaks=seq(0,4,by=0.25))) %>% 
  select(ETIQUETA,Year,Rho)
canarias <- canarias %>% full_join(rates) 

# The colour palette is set so that rates lower than 1 are painted green 
# and from 1 onwards they become increasingly red:
coloresMapa <- colorRampPalette(brewer.pal(11,"RdYlGn"))(11)[c(10:8,6:1)]
# Map:
ggplot() +
  layer_spatial(canarias, aes(fill = Rho), col = 'black', show.legend = TRUE) +
  coord_sf(crs=sf::st_crs(4326)) +
  labs(x = NULL,  y = NULL, fill=expression(rho)) +
  scale_fill_manual(values=coloresMapa) +
#  scale_fill_brewer(palette="RdYlGn", direction=-1) +
  theme(
    legend.position = "right",#c(0.9, 0.85),
    legend.text.align = 0,
    legend.background = element_rect(fill = alpha('white', 0.0)),
    legend.text = element_text(size = 12, hjust = 0, color = "#4e4d47"),
    legend.title = element_text(size = 14),
    plot.title = element_text(size = 18, hjust = 0, color = "#4e4d47"),
    plot.subtitle = element_text(size = 20, hjust = 0.8, face = "italic", color = "#4e4d47"),
    plot.caption = element_text(size = 9, hjust = 0, color = "#4e4d47"),
    plot.margin = unit(c(.5,.5,.2,.5), "cm"),
    panel.border = element_blank(),
    axis.title=element_blank(),
    axis.text=element_blank(),
    axis.ticks=element_blank(),
    panel.grid.major = element_blank(),
    panel.background = element_rect(fill = "azure")
  ) +
  facet_wrap(~Year,ncol=1) +
  theme(strip.text = element_text(size = rel(1.2)))
ggsave("figure4.eps",dpi=1200,width=238,height=180,units = "mm")
```

### Figure 5: Evolution of infant mortality in Canary Islands and Spain
```{r}
ims <- read.xlsx("IM_España (INE).xlsx")
ims2 <- ims %>% 
  select(Year=Año, Spain=Total.Nacional, `Canary Islands`= `05.Canarias`) %>% 
  mutate(Year=as.numeric(Year))
ims2 %>% 
  filter(Year>=1990&Year<2021) %>% 
  pivot_longer(-Year,names_to = "Region",values_to = "IM") %>% 
  ggplot(aes(x=Year,y=IM,linetype=Region, group=Region))+
  geom_line()+
 # theme_bw() +
  labs(y="Infant Mortality",color="") +
  scale_x_continuous(breaks=seq(1990,2020,by=5)) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave(file="figure5.eps",width=119,height=119,units="mm",dpi=1200)
```


### Frequentist estimation
```{r}
library(lme4);
mm <- glmer(Death ~   Period  + (1 | County),offset = log(Birth), data=dt,family=poisson(link="log"))
summary(mm)
```
